XRONOS: an open data infrastructure for archaeological chronology

### 2024-11-29 paper.qmd
added some examples for kintighs call

XRONOS is our answer to Kintigh's call [@Kintigh2006] for digital infrastructures that don't just provide access to chronological data but enables researchers to "archive, access, integrate, and mine disparate data sets".
It parallels and draws inspiration from several similar initiatives within and outwith archaeology, such as <!-- TODO examples -->.
To improve upon existing global syntheses of radiocarbon dates (see @sec-global-compilations), we aimed to develop a living infrastructure that both continually collected data from diverse sources and presented a seamless single database to the user.
The Global Biodiversity Information Facility [GBIF, <https://gbif.org>, @CanhosEtAl2004]—which provides a single, consistent interface to many sources of global biodiversity data—has served as an exemplar for us in this regard.

>>>>

XRONOS is our answer to Kintigh's call [@Kintigh2006] for digital infrastructures that don't just provide access to chronological data but enables researchers to "archive, access, integrate, and mine disparate data sets".
It parallels and draws inspiration from several similar initiatives within and outwith archaeology, such as <!-- TODO examples
Examples:
https://link.springer.com/article/10.1007/s10816-018-9399-6
https://link.springer.com/article/10.1007/s10816-010-9098-4
https://www.cambridge.org/core/journals/latin-american-antiquity/article/public-database-of-archaeological-resources-on-easter-island-rapa-nui-using-google-earth/F30677F6FC99B4C2BC762BCC98FD0966
https://www.cambridge.org/core/journals/advances-in-archaeological-practice/article/data-integration-in-the-service-of-synthetic-research/E297C57441239AB35053F987AA758EB9
https://www.cambridge.org/core/journals/advances-in-archaeological-practice/article/tdar/FCE1056A210A452E5FE65560763D4CC7
https://www.cambridge.org/core/journals/advances-in-archaeological-practice/article/googlebased-freeware-solution-for-archaeological-field-survey-and-onsite-artifact-analysis/7914F673D1E1CAD96E5270E794CDFFD9
https://link.springer.com/article/10.1007/s10816-015-9272-9
https://link.springer.com/article/10.1007/s10816-015-9251-1
https://link.springer.com/article/10.1007/s10816-015-9240-4
https://anatomypubs.onlinelibrary.wiley.com/doi/10.1002/ar.23130
https://academic.oup.com/dsh/article/29/3/349/2938128
https://www.sciencedirect.com/science/article/pii/S0167739X13000678?via%3Dihub

-->.
To improve upon existing global syntheses of radiocarbon dates (see @sec-global-compilations), we aimed to develop a living infrastructure that both continually collected data from diverse sources and presented a seamless single database to the user.
The Global Biodiversity Information Facility [GBIF, <https://gbif.org>, @CanhosEtAl2004]—which provides a single, consistent interface to many sources of global biodiversity data—has served as an exemplar for us in this regard.
